Gap formation and stability in non-isothermal protoplanetary discs

### 2014-11-18 paper/results.tex
begin second edits - MKL

\section{Growth of non-axisymmetric modes without the influence of the
  planet}\label{linear1}
In this section, the planet is introduced at $t=20P_0$ and 
its potential switched on over $10P_0$. At $t=30P_0$ we switch off the
planet potential and azimuthally average the surface density, energy
and velocity fields. At this point the planet has carved a partial
gap and the RWI has not yet occured.
 We then perturb the surface density for $r>r_p$ and continue to 
evolve the disc. We impose  sinusoidal perturbations with 
azimuthal wavenumbers $m\in[1,10]$ and random amplitudes within $\pm 0.01$
 times the local surface density.% {\bf what are PERTMIN and PERTMAX values?}  
This procedure allows us to analyse the growth of 
non-axisymmetric modes associated with the gap, but without
complications from non-axisymmetry arising directly from disc-planet
interaction.

@@

Simulations here employ a resolution of $(N_r,N_{\phi})=(1024,2048)$
with an open boundry condition at the outer disk edge $r_\mathrm{out}=25r_\mathrm{in}$. We consider 
%for a time up to $200P_0$ allowing high resolution and detection of high
%$m$ structures.
cases of $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
and slowly cooled discs. % {\bf boundary condition for these sims?}

\subsection{Gap structure}
%Since stability of the gap is of interest 
We first examine the gap structure formed by planet-disc
interaction as a function of the cooling time. The azimuthally-averaged 
gap profiles are shown in Fig. \ref{intial1D} for varying
$\tilde\beta$. Gaps formed with lower $\tilde\beta$ (faster cooling)
are deeper with steeper gradients at the gap edges. Faster cooling rates also 
increase and decrease the surface density maxima and 
minima increase, respectively. However, a clean gap does not form
in this short time period.

@@

Larger $\tilde\beta$ values result in higher disc aspect ratios $h=H/r$,
i.e. higher temperatures. Heating mostly occur at the gap edges
due to planet-induced spiral shocks. Decreasing the cooling time
implies that this heat is retained in the disc. In the inviscid limit the gap
opening condition is $r_h\gtrsim H$ or $q\gtrsim 3h^3$
\citep{crida06}. 
% which is a value directly correlated to temperature in our model. 
%For a planet to carve a gap in the surrounding material it has been
%shown that $q>H^3$ is needed \citep{crida06}. 
This indicates that for hotter discs (higher
$h$), it becomes more difficult for a planet of fixed $q$ to open a
gap. This explains the shallower gaps in surface density when
$\tilde{\beta}$ is increased.

@@

The important consequence of a heated gap edge is that the
generalized vortensity profiles become smoother with increasing
cooling times, with the extrema becoming less pronounced. Previous locally
isothermal disc-planet simulations show the RWI associated with PV
minima \citep{li05,lin10}. We can therefore expect the RWI to be associated with
minima in the generalized vortensity (corresponding to local surface
density maxima) in the non-isothermal case. Because the extrema are
less sharp, the RWI is expected to be weaker and the gap to be more
stable with longer cooling times.

@@

{\bf
\subsection{Axisymmetric stability}
Check that the gap profiles are stable against axisymmetric
stability. Can use analytical criteria given in section 3.1 of Li et
al (2000, ApJ, 533, 1023). No plot needed, just statements. 
}

The intial planet-disk interaction form bumps and grooves in the gap profiles
 which can potentially be unstable due to axisymmetric instabilities. The
 generalized local axisymmetric stability condition is the Solberg-Hoiland
 criterion,
\begin{equation}
 \kappa^2+N^2 \geq 0 $ where $ N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r} \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right)
\end{equation}
where $N$ is the Brunt-V\"ais\"al\"a frequency for the disk due to entropy
 variation which is non-zero for non-adiabatic models such as being considered.
 At  $t=30P_0$ the outer gap edge of $r=2.5r_h$,  where the RWI is excited the
 criterion value, is found to be approximately $0.002$ for all $\tilde\beta$
 which is the local minimum in the gap region which indicates local
 axisymmetric stability of gap. The Solberg-Hoiland criteria is similary
 satisfied for the entire disk and for all times within the simulation length
 with little change of values. Thus for all cases of $\tilde\beta$ the
 planet-induced gaps formed are stable to axisymmertic instabilities.

@@

In all cases we observe growth of non-axisymmetric structures after the
planet potential is switched off and the system subject to random
perturbations. An example is shown in Fig. \ref{linearmodes} for 
$\tilde{\beta}=10$. We charcterize these
modes with an azimuthal wavenumber $m$ and growth rate $\gamma(m)$ as defined by
Eq.~\ref{fouriertransform}---\ref{growth} with radial range of $[2r_h,5r_h]$. Table \ref{modetable}
lists the growth rates for the three cooling
times, measured during linear growth.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
    0cm 0cm 0cm]{figures/linear_stability}
  \caption{Evolution of azimuthal Fourier modes of disc surface
    density, non-dimenionlized by the axisymmetric background keplerian mode
 $\Sigma_0(t=0)$ for the
    `planet-off' simulations with $\tilde{\beta}=10$. Colours correspond
    to different $m$ values. The $m=3$ component is the fastest growing
    mode during linear growth and has a corresponding
    $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
% {\bf is the $\Sigma_0$ at t=0?} }  
\end{figure}

@@

Table \ref{modetable} show that as
$\tilde{\beta}$ was increased from $ 0.1\rightarrow10$ the dominant
azimuthal Fourier mode decreased from $ m=9\rightarrow3$ and the
respective growth rate decreased from $ \gamma/\Omega(r_p)=0.079
\rightarrow 0.017$. However, despite two orders of magnitude increase in the
cooling time, the instability remains dynamical with growth time
$\lesssim 10P_0$. Snapshots of the instability in $r-\phi$ plane for
the different $\tilde\beta$ are shown in Fig \ref{2Dlinear}.

@@

\begin{figure}
  \centering
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity50}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity67}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity75}
  }  
  \caption{Generalized vortensity perturbation (relative to $t=0$) for
    cases of $\tilde{\beta}=0.1,1,10$ (left,middle,right) during
    the growth of non-axisymmetric modes. The planet potential has
    been switched off.  The number of vortices
    decrease as $\tilde{\beta}$ increases. Note that snapshots are
    taken later because it takes longer for the vortices to grow and
    become visible with increasing cooling time. 
    %    correspondingly increase to capture the longer growth phases of
    %    the instabilities. 
    \label{2Dlinear} {\bf I'm curious to know if plots clearer if
      perturbations are measured relative to t=30? remember t=30 is
      our reference state for instability. But we'll probably keep
      these for consistency (later non-linear plots are all measure
      with respect to t=0).}} 
\end{figure}

\subsubsection{Nearly adiabatic discs}
\label{adiabatic_section}
%{\bf a case with extrmely long cooling time (or adiabatic run) in
%  order to look at effect of heated gap edge.  prelim result: not
%  important, at t=30, gap edge only heats to $H/r\simeq0.06$ even for
%  purely adiabatic disk. change in $h$ not important for linear
%  perturbations (but important for setting up the basic state). 
%}
The above `planet-off' simulations are not formally linear
stability calculations, because the cooling time is shorter
than the instability growth time, 
%
%as we do not capture the effects of a heated
%disk on the instability growth. 
$t_c<\gamma^{-1}$.  
Thus the disk cools back to its intial value of $h=0.05$
before or during the instability growth, implying we do not have a
steady basic state to formulate a standard linear stability
problem. %This means the disc temperture of $h=0.05$ in the the above cases

@@

According to \cite{li00}, increasing $h$ increases linear growth rates
of the RWI because it is pressure-driven. However, in the case 
of disc-planet interaction, increasing $h$ has a stabilizing effect
through the gap profile because it results in smoother gap
edges. The fact that we observe smaller growth rates as $h$ is
increased indicate that for planetary gaps, the importance of $h$ on
the \emph{linear} RWI is through the setting the gap profile (i.e. basic
state for the instability).

@@

@@

We now consider long term simulations of gap-edge vortices for
$\tilde{\beta}=0.1,0.5,1,5,10$. The planet potential is kept on
throughout.  
% up to a total of
%$2000P_0$. 
%Planets were left free to interact dynamiclly with the disk
%and vortices after $t=30P_0$ as apposed to previous section.  
We employ a grid with $(N_r,N_{\phi})=(512,1024)$ 
in order for these simulations to be computationally feasible. We also use a larger
disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimize boundary
effects on vortex evolution. We apply an open boundary at
$r_\mathrm{out}$.

Preliminary simulations with same numerical scheme but lower resolution of
 $(N_r,N_{\phi})=(256,512)$ where also done for the above $\tilde\beta$ values
with the additional inclusions of $\tilde{\beta}=0.25,0.75,2.5,7.5$.
Vortex evolution and behaviour are convergent with
higher resolution simulations as discussed in this section.
 {\bf mention lower resolution preliminary runs - 
similar results to those presented here (?). want to demonstrate convergence}

@@

\subsection{Generic evolution}

The linear growth of the RWI and vortex-formation is followed by 
vortex merging. Merging timescales are independent of $\tilde\beta$ value,
 and by $60P_o$ only one vortex remains as the growth of these modes are now
 accelerated by planetary influence as apposed to `planet off' simulations.
%i.e. only the $m=1$ mode remains.

The evolution of the amplitude of the $m=1$ surface density mode
 over the radial range of $[2r_h,10r_h]$ is shown in Fig~\ref{lifetimeplot}
 for different $\tilde\beta$.
% {\bf should state (also for planet-off runs)
 % what the radial range is used for averaging the Fourier amplitudes}

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
    0cm 0cm 1cm]{figures/longterm_stability}
  \caption{Log plot of the $m=1$ mode non-dimenionlized by the $m=0$
    background mode for longterm non-linear simulations corrsponding
    to vortex lifetime. The longest living mode correpsonds to
    $\tilde\beta=5.0$ (the red line). \label{lifetimeplot}} 
\end{figure}

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex grows in intensity. As the vortex forms it has a
 value of $Ro\approx-0.1$ for all $\tilde\beta$
 indicating anti-cyclonic
motion. As the $m=1$ amplitude grows the relative spin of the vortex
continuously grows along with it, reaching values of
$Ro\approx-0.4$ characteristicly for all cooling rates.

Surface density pertibations relative to the intial disk
at the center of vortices can reach values of $\Delta\Sigma/\Sigma_0=7$ 
 for all cases of $\tilde\beta$ in this quasi-steady state and
can be found as high as $\Delta\Sigma/\Sigma_0= 9$ times for the 
$\tilde\beta=10$ case.
 The gap profiles formed also apply a surface density boost in the vortex 
region as the vortices are located at the gap edge maxima.
 The average value of relative density pertibation in the outer gap region of 
$r=r_{p}+2.5r_h$
has a value $\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$ for all times and
 cooling rates.
Thus the surface density increases caused by gap formation are small 
compared to the max amplitudes of the vortices.
 Vortex growth is attributed to
the continuous generation of vorticity by the planet-disc interactions. 
This is supported by the evidence that in long term simulations where the
planet is removed, such as seen in Fig \ref{planetofflifetimeplot},
 the amplitude of
 the vortices are not seen to grow without the planet influence.

{\bf is vortex merging affected by $\tilde{\beta}$? e.g. timescale
  taken to reach one vortex as a function of cooling}
{\bf for what cooling are these values quoted for? do
  these values depend on cooling?}
 {\bf is the evidence for this the fact that growth is not observed in
  planet-off simulations?}

{\bf note that there
  are two origins for the surface density boost relative to $t=0$: 1) gap
  formation, which increases the surface density at the location of
  the instability (since they happen at surface density max., and 2)
  vortex instability). what is the surface density boost relative to
  the gap profile?}

@@

\subsection{Non-linear instability without planet influence} \label{nonlinearplanetoff}
Non-linear observations were also made for the `planet-off' simulations.
After the linear growth phase of the vortices, vortex merging takes hold but
 now on timescales of up to $150P_0$ as the vortices grow as linear
 instabilities discussed previously. The vortex merging time is depedendent on
 the growth rates of the modes and saturation timescales with the
slowest growing modes in $\tilde\beta=10$ taking the longest. Similar
 quasi-steady vortices form as in the cases with continuous planet interaction
but the long-term behaviour is significantly
 different without the planet as seen in Fig.\ref{planetofflifetimeplot}.
The quasi-steady vortices formed have significantly lower density pertibations
and now slowly decrease in amplitute in time as apposed to previously
where their growth was supported by planet interactions. The vortices simply
dissipate viscously with dissipation rate decreasing with larger $\tilde\beta$.
The vortices are never found to undergo the drastic sudden dissipation effects
that characterize the planet induced growing cases do.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
    0cm 0cm 1.1cm]{figures/longterm_planetoff}
  \caption{Similar plot as Fig. \ref{lifetimeplot} but with `planet-off' 
simulations. Vortices dissipate viscously without planet driving influence} \label{planetofflifetimeplot}
\end{figure}

>>>>

\section{Growth of non-axisymmetric modes without the influence of the
  planet}\label{linear1}
In this section, the planet is introduced at $t=20P_0$ and 
its potential switched on over $10P_0$. At $t=30P_0$ we switch off the
planet potential and azimuthally average the surface density, energy
and velocity fields. At this point the planet has carved a partial
gap and the RWI has not yet occured.
 We then perturb the surface density for $r>r_p$ and continue to 
evolve the disc. We impose  sinusoidal perturbations with 
azimuthal wavenumbers $m\in[1,10]$ and random amplitudes within $\pm 0.01$
 times the local surface density.% {\bf what are PERTMIN and PERTMAX values?}  
This procedure allows us to analyse the growth of 
non-axisymmetric modes associated with the gap, but without
complications from non-axisymmetry arising directly from disc-planet
interaction (i.e. planet-induced wakes).

@@

Simulations here employ a resolution of $(N_r,N_{\phi})=(1024,2048)$
with open boundaries at $r=r_\mathrm{in}$ and
$r_\mathrm{out}=25r_\mathrm{in}$. We consider   
%for a time up to $200P_0$ allowing high resolution and detection of high
%$m$ structures.
cases of $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
and slowly cooled discs. % {\bf boundary condition for these sims?}

\subsection{Gap structure}
%Since stability of the gap is of interest 
We first examine the gap structure formed by planet-disc
interaction as a function of the cooling time. The azimuthally-averaged 
gap profiles are shown in Fig. \ref{intial1D} for varying
$\tilde\beta$. Gaps formed with lower $\tilde\beta$ (faster cooling)
are deeper with steeper gradients at the gap edges. Faster cooling rates also 
increase (decrease) the surface density maxima (minima). However, a
clean gap does not form in this short time period.

@@

Larger $\tilde\beta$ values result in higher disc aspect ratios $h=H/r$,
i.e. higher temperatures. Heating mostly occur at the gap edges
due to planet-induced spiral shocks. Decreasing the cooling time
implies that this heat is retained in the disc. In the inviscid limit the gap
opening condition is $r_h\gtrsim H$ or $q\gtrsim 3h^3$
\citep{crida06}. 
% which is a value directly correlated to temperature in our model. 
%For a planet to carve a gap in the surrounding material it has been
%shown that $q>H^3$ is needed \citep{crida06}. 
This indicates that for hotter discs (higher
$h$), it becomes more difficult for a planet of fixed $q$ to open a
gap. This explains the shallower gaps in surface density when
$\tilde{\beta}$ is increased. 
%However, notice the gap structure
%changes most significantly for $\beta =0.1 \to 1.0$. For $\beta \geq
%1$

@@

The important consequence of a heated gap edge is that the
generalised vortensity profiles, $\tilde{\eta}$, becomes smoother with increasing
cooling times, with the extrema becoming less pronounced. Previous locally
isothermal disc-planet simulations show the RWI associated with PV
minima \citep{li05,lin10}. We can therefore expect the RWI to be associated with
minima in the generalised vortensity (corresponding to local surface
density maxima) in the non-isothermal case. Because the extrema are
less sharp, the RWI is expected to be weaker and the gap to be more
stable with longer cooling times.

@@

\subsection{Axisymmetric stability}
%Check that the gap profiles are stable against axisymmetric
%stability. Can use analytical criteria given in section 3.1 of Li et
%al (2000, ApJ, 533, 1023). No plot needed, just statements. 
%}
The intial planet-disc interaction form bumps and grooves in the gap profiles
which can potentially be unstable due to axisymmetric instabilities. The
generalised local axisymmetric stability condition is the Solberg-Hoiland
criterion,  
\begin{align}
  \kappa^2+N^2 \geq 0 
\end{align}
where
\begin{align}
 N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r}
 \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial
     r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right) 
\end{align}
is the square of the Brunt-V\"ais\"al\"a frequency.  
%for the disk due to entropy
%variation which is non-zero for non-adiabatic models such as being
%considered. 
At the outer gap edge $r=r_p+2.5r_h$,  where the RWI is excited
(see below), we find $\kappa^2 + N^2$ reaches local minimum with a value
$\sim 0.002$ (code units) for all $\tilde\beta$. The
Brunt-V\"ais\"al\"a frequency is $N\sim 10^{-5}$. The Solberg-Hoiland
criteria is similary satisfied for the entire disc throughout the
simulations. % with little change of values. 
Thus for all values of $\tilde\beta$ the planet-induced gaps are
stable to axisymmertic instabilities.

@@

We now examine the evolution of the gap for $t>30P_0$, with the
planet potential switched off, but with an added surface density
perturbation. For all three cooling times $\tilde{\beta}=0.1,\, 1.0,\,
10$, we observe exponential growth of non-axisymmetric
structures. % after the 
%planet potential is switched off and the system subject to random
%perturbations.  
An example is shown in Fig. \ref{linearmodes} for 
$\tilde{\beta}=10$. We charcterize these
modes with an azimuthal wavenumber $m$ and growth rate $\gamma(m)$ as defined by
Eq.~\ref{fouriertransform}---\ref{growth}. Mode amplitudes were
averaged over $r-r_p\in[2,5]r_h$. Table \ref{modetable}
lists the growth rates measured during 
linear growth for 5 values of $m$ centred around the maximum.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
  0cm 0cm 0cm]{figures/linear_stability}
  \caption{Evolution of azimuthal Fourier modes of disc surface
    density, non-dimenionlized by the initial axisymmetric background mode 
    $\Sigma_0(t=0)$ for the
    `planet-off' simulations with $\tilde{\beta}=10$. Colours correspond
    to different $m$ values. The $m=3$ component is the fastest growing
    mode during linear growth and has a corresponding
    $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
  % {\bf is the $\Sigma_0$ at t=0?} }  
\end{figure}

@@

Table \ref{modetable} show that as
$\tilde{\beta}$ is increases from $ 0.1\rightarrow10$ the dominant
azimuthal Fourier mode decreases from $ m=9\rightarrow3$ and the
respective growth rate decreases from $ \gamma/\Omega(r_p)=0.079
\rightarrow 0.017$. However, despite two orders of magnitude increase in the
cooling time, the instability remains dynamical with characteristic  growth time
$\lesssim 10P_0$. Snapshots of the instability in for  
the different $\tilde\beta$ are shown in Fig \ref{2Dlinear}.

@@

\begin{figure}
  \centering
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity50}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity67}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity75}
  }  
  \caption{Generalized vortensity perturbation (relative to $t=0$) for
    cases of $\tilde{\beta}=0.1,1,10$ (left,middle,right) during
    the growth of non-axisymmetric modes. The planet potential has
    been switched off.  The number of vortices
    decrease as $\tilde{\beta}$ increases. Note that snapshots are
    taken later for increasing $\tilde{\beta}$ because it takes longer
    for the vortices to grow and become visible with increasing cooling time. 
    % correspondingly increase to capture the longer growth phases of
    % the instabilities. 
    \label{2Dlinear} 
%{\bf I'm curious to know if plots clearer if
%      perturbations are measured relative to t=30? remember t=30 is
%      our reference state for instability. But we'll probably keep
%      these for consistency (later non-linear plots are all measure
%      with respect to t=0).}
  } 
\end{figure}

\subsubsection{Nearly adiabatic discs}
\label{adiabatic_section}
%{\bf a case with extrmely long cooling time (or adiabatic run) in
%  order to look at effect of heated gap edge.  prelim result: not
%  important, at t=30, gap edge only heats to $H/r\simeq0.06$ even for
%  purely adiabatic disk. change in $h$ not important for linear
%  perturbations (but important for setting up the basic state). 
%}
The above `planet-off' simulations are not formally linear
stability calculations, because the cooling time is shorter
than the instability growth time, 
%
%as we do not capture the effects of a heated
%disk on the instability growth. 
$t_c<\gamma^{-1}$.  
Thus the disk cools back to its intial temperature corresponding to
$h=0.05$ before or during the instability growth, so we do not
have a steady basic state to formulate a standard linear stability 
problem. %This means the disc temperture of $h=0.05$ in the the above cases

@@

According to \cite{li00}, increasing $h$ increases linear growth rates
of the RWI because it is pressure-driven. However, in the case 
of disc-planet interaction, increasing $h$ has a stabilizing effect
through the gap profile because it results in smoother gap
edges. The fact that we observe smaller growth rates as $h$ is
increased indicates that for planetary gaps, the importance of $h$ on
the \emph{linear} RWI is through the setting the gap profile, i.e. basic
state for the instability (as opposed to the linear response).

@@

\subsection{Non-linear evolution without planet
  influence} \label{nonlinearplanetoff} 
{\bf typical rossby numbers for these `planet-off' vortices? do
  stronger vortices (beta=0.1) dissipate faster? if we want to say the
  vortices dissipate viscously, then need to quote a viscosity
  (e.g. reynolds stress) and compute a viscous time and compare to the
  observed decay time. Or attribute to numerical viscosity. It might
  be simplest to just say the vortices decay without the planet. This
  section is mainly used to contrast with planet-on simulations, so
  don't need to go into detail. 
}

We also extended these `planet-off' simulations well into the non-linear
regime. After the linear growth phase of the vortices, vortex merging
takes hold on timescales of up to $150P_0$, until there is one vortex
left. We find the vortex merging
time is depedendent on the growth rates of the modes and saturation
timescales, with the slowest growing modes in $\tilde\beta=10$ taking
the longest to merge.

Fig.\ref{planetofflifetimeplot} shows evolution of the $m=1$ surface
density amplitude, which represents the post-merger single vortex. For
completeness we also ran intermediate cases with $\tilde{\beta}=0.5$
and $5.0$. The vortices simply decay, on a timescale of $O(10^3)$
orbits. We will see in the that section that this behaviour is very
different to when the planet potential is kept on. Interestingly,
though, vortices arising from the stronger instabilities (shorter
cooling times or lower $\tilde{\beta}$) decays faster.

%Similar
%quasi-steady vortices form as in the cases with continuous planet interaction
%but the long-term behaviour is significantly
%different without the planet as seen in 
%The quasi-steady vortices formed have significantly lower density pertibations
%and now slowly decrease in amplitute in time as apposed to previously
%where their growth was supported by planet interactions. 
%The vortices simply
%dissipate viscously with dissipation rate decreasing with larger $\tilde\beta$.

%Non-linear observations were also made for the `planet-off'
%simulations.
% as the vortices grow as linear
% instabilities discussed previously.

%The vortices are never found to undergo the drastic sudden dissipation effects
%that characterize the planet induced growing cases do.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1.1cm]{figures/longterm_planetoff}
  \caption{Long term simulations without the planet potential after
    the gap is set up. The $m=1$ surface density component at the
    outer gap edge is shown.
  } \label{planetofflifetimeplot}
\end{figure}

@@

We now examine long-term simulations of gap-edge vortices for
$\tilde{\beta}=0.1,0.5,1,5,10$. The planet potential is kept on
throughout.  
% up to a total of
%$2000P_0$. 
%Planets were left free to interact dynamiclly with the disk
%and vortices after $t=30P_0$ as apposed to previous section.  
We employ a grid with $(N_r,N_{\phi})=(512,1024)$ in order for these
simulations to be computationally feasible. We also use a larger
disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimize boundary
effects on vortex evolution, and apply an open boundaries at
$r=r_\mathrm{in},\,r_\mathrm{out}$. We comment that lower-resolution
simulations with $(N_r,N_{\phi})=(256,512)$ were initially carried
out, which show similar behaviour and trends as the high-resolution
runs reported below.

%where also done for the above $\tilde\beta$ values
%with the additional inclusions of $\tilde{\beta}=0.25,0.75,2.5,7.5$.
%Vortex evolution and behaviour are convergent with
%higher resolution simulations as discussed in this section.
% {\bf mention lower resolution preliminary runs - 
%similar results to those presented here (?). want to demonstrate convergence}

@@

\subsection{Generic evolution} 
The linear growth of the RWI and vortex-formation is followed by 
vortex merging. We now find merging timescales independent of
$\tilde\beta$, and by $60P_o$ only one vortex remains. 
%as the growth of these modes are now
%accelerated by planetary influence as apposed to `planet off' simulations.
% i.e. only the $m=1$ mode remains. 
The evolution of the amplitude of the $m=1$ surface density mode,
averaged over $r-r_p\in[2,10]r_h$ is shown in Fig~\ref{lifetimeplot} 
for different $\tilde\beta$. 
{\bf any particular reason for the larger radial range for average?
  did we get bigger vortices?} 
% {\bf should state (also for planet-off runs)
% what the radial range is used for averaging the Fourier amplitudes}

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1cm]{figures/longterm_stability}
  \caption{Evolution of the $m=1$ surface density component,
    non-dimenionlised by the initial $m=0$ component, for long term
    simulations with the planet potential kept on. The case with 
    $\tilde\beta=5.0$ (red) gave the longest-living vortex. {\bf
      remember to update to 
      include vortex dissipation for all cases} \label{lifetimeplot}} 
\end{figure}

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex intensifies. As the vortex forms it has a
Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
indicating anti-cyclonic
motion. As the $m=1$ amplitude grows the Rossby number also increases
in magnitude, 
%relative spin of the vortex
%continuously grows along with it
, reaching a characteristic value of $Ro\approx-0.4$ for all cooling
rates. {\bf is the the value reached just before decay for all cooling?}

Surface density pertibations relative to $t=0$
at the center of vortices can reach values of $\Delta\Sigma/\Sigma_0=7$ 
for all cases of $\tilde\beta$ in this quasi-steady state and
can be found as high as $\Delta\Sigma/\Sigma_0= 9$ times for the 
$\tilde\beta=10$ case. 
The gap profiles formed also apply a surface density boost in the vortex 
region as the vortices are located at the gap edge maxima.
The average value of relative density pertibation in the outer gap region of 
$r=r_{p}+2.5r_h$
has a value $\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$ for all times and
cooling rates.
Thus the surface density increases caused by gap formation are small 
compared to the max amplitudes of the vortices.
Vortex growth is attributed to
the continuous generation of vorticity by the planet-disc interactions. 
This is supported by the evidence that in long term simulations where the
planet is removed, such as seen in Fig \ref{planetofflifetimeplot},
the amplitude of
the vortices are not seen to grow without the planet influence.

%{\bf is vortex merging affected by $\tilde{\beta}$? e.g. timescale
%  taken to reach one vortex as a function of cooling}
%{\bf for what cooling are these values quoted for? do
%  these values depend on cooling?}
% {\bf is the evidence for this the fact that growth is not observed in
%  planet-off simulations?}   
%{\bf note that there
%  are two origins for the surface density boost relative to $t=0$: 1) gap
%  formation, which increases the surface density at the location of
%  the instability (since they happen at surface density max., and 2)
%  vortex instability). what is the surface density boost relative to
%  the gap profile?}

@@
